The Talking Heads experiment

### 2015-04-28 chap10/chap10.tex
format relevant papers section

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.50\textwidth]{chap10/figs/look-ball.pdf}
\includegraphics[width=.45\textwidth]{chap10/figs/views-ball.pdf}}
\caption{\label{f:look-ball}Left: Fr\'{e}d\'{e}ric Kaplan interacts with the \textsc{aibo} robot in an experiment for the social learning of lexicons.
Right: View of the ball through the eye of the robot. As the same object is now seen from many different viewpoints it 
becomes much more challenging to consistently recognise it.}
\end{figure}
The first \textsc{aibo} experiments did not involve a population of agents but rather a single agent that was interacting directly with a human
experimenter. The agent had to learn the lexicon of the human through social 
interaction within the context of situated language games. The 2001
paper ``\textsc{aibo}'s first words'' by Steels and Kaplan \nocite{Steels:01h} described experiments in which an \textsc{aibo} robot 
learned words about objects in the environment, such as `ball' or `smiley', as well as words for actions, such as 
`sit down' or `stand up'.

A typical example of a game is the DO game, used to learn the names of behaviours.\is{DO game} 
The robot is programmed with a repertoire of behaviours, such as 
walk, sit down, stand up, look left/right/up/down, push ball, etc. When the human utters a word, the hearing robot 
looks up the word in its own memory and performs that behaviour. If it does not know the word, it 
performs one of the behaviours for which it does not have a word yet. 
If this behaviour is the wrong one, the robot receives feedback and correction, 
thus learning (i) that the behaviour is not associated with the word and (ii) what the right word is for the behaviour 
just performed. When the right behaviour was chosen, the association between this behaviour 
and the word can be stored if the association was not yet in memory, or reinforced if it was using the 
lateral inhibition learning strategy. 
Games therefore always come with two variant scripts. There is one with a successful interaction, which then 
leads to a reinforcement of the existing lexicon, and one with a non-successful interaction, where the 
robot gets corrected and learns a new word. Here are the two variants for the DO Game: 
\begin{center}
\begin{tabular}{ l  l }
\lsptoprule
{\itshape Reinforcement script}&{\itshape Correction script} \\ \midrule
Human: Listen, Walk. & H: Listen, Walk.  \\ 
R: Walk? & R: Walk? \\
 {\itshape (walks)} & {\itshape (sits)} \\
H: Yes & H: No, this is `sit'. \\
& R: Sit ? \\ 
& H: Yes. \\ 
\lspbottomrule
\end{tabular}
\end{center}
The scripts of the agents are represented internally using probabilistic finite state machines with 
temporal annotations. This computational formalism (pioneered by Rodney Brooks) was also used to program the 
physical behaviour of the robots (\figref{fig:fsm}).\footnote{
This architecture was first described in \cite{Brooks:1986}. 
For the experiments I implemented a LISP-based Behaviour Language that ran also on the \textsc{aibo}.}

@@

The \textsc{aibo} experiments were the first robotic experiments demonstrating social, situated learning in action. 
They were highly relevant for discovering the subtle interaction patterns that would work and 
could be realistically implemented in human-robot scenarios. The goal of these experiments was 
not to address the question of the origins of a 
new language in a population of agents. However, another experiment, still with the \textsc{aibo} robots,  
did substantially advance the state of the art: the Perspective Reversal Experiment carried out by Martin Loetzsch 
and Luc Steels in 2004--2005.\footnote{
Very little was published about this experiment, partly because reviewers had the greatest difficulty to grasp 
its underlying rationale or understand the technical details. The main paper is \cite{Steels:2008spatial}. 
Subsequent work by Michael Spranger significantly scaled up the experiment, as discussed in the next chapter.}

\subsection{The Perspective Reversal Experiment}

The Perspective Reversal Experiment \is{Perspective Reversal} focused on the question how a spatial language could emerge that employed
perspective reversal. It was the first experiment to systematically compare different language strategies, thus 
introducing for the first time the {\itshape comparative method} \is{comparative method} 
that would inform all later language game experiments.

@@

The Perspective Reversal Experiment applied this methodology to explain a remarkable universal feature 
of human languages, namely, 
that speakers may use a different perspective on the scene than their own when 
conceptualising what to say, and that they may explicitly mark perspective switching, using lexical or grammatical means. Suppose
a speaker is facing a hearer, then it indeed makes a big difference whether she says 
\emph{the door to your left}  or \emph{the door to my left}. If the speaker says
\emph{the door to my left} she expects the hearer to perform an egocentric
perspective transformation and see the situation from her own point of view. Perspective switching is here signalled by 
explicitly expressing which perspective is used: {\itshape your} left versus {\itshape my} left. 
The German prepositions \emph{herein} and \emph{hinein}, both meaning `into' are another example of perspective marking. They
distinguish whether the direction of movement is 
towards the speaker \emph{herein} or away from the speaker \emph{hinein}, analogous to English \emph{come} (to where I am) 
versus \emph{go} (to another location). Although there are obvious differences in how languages express
perspective, there can be no doubt that perspective marking is pervasive across 
languages and language families, and that speakers make abundant use of it.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.65\textwidth]{chap10/figs/tokyo-persp.pdf}}
\caption{\label{fig:tokyo-persp}Luc Steels demonstrates live the Perspective Reversal Experiment in Tokyo, April 2005. \textsc{aibo} robots track 
the movement of a ball and 
the robot acting as speaker describes to the hearer an observed movement, possibly taking into account the 
perspective of the hearer. For example, the speaker might say `the ball rolls from my left to your right'.}
\end{figure}

\enlargethispage{1\baselineskip}
The Perspective Reversal Experiment worked as follows. A population of \textsc{aibo} robots roam around
freely in an unconstrained in-door environment. As soon as one sees the
ball, it comes to a stop and searches for another robot nearby, which also
looks for the ball and stops when it sees it. Then the human
experimenter pushes the ball with a stick so that it rolls a short
distance (see \figref{fig:tokyo-persp}). The two robots next play a description game. One robot (acting as the `speaker') 
describes the ball-moving event to the other robot (the ``hearer''), going through the same semiotic cycle as the Talking 
Heads agents. The game is a success if the meaning conveyed by the speaker is compatible with the hearer's 
own perception of the scene.

@@

Within this experimental setting three different strategies were tried in a long-term experiment where
a new strategy was injected at certain stages and performance monitored: 
\begin{itemize}
\item In stage \textsc{i}, agents use 
the Discrimination Game strategy and the Naming Game strategy discussed in chapters 4 and 5. A 
lexicon is emerging and there are some successful games (those for which the speaker and the hearer see the situation from 
the same perspective). But the lexicon keeps expanding and success is very limited as games where robots have a different 
perspective on the rolling ball all fail. 
\item In stage \textsc{ii}, agents have an extended 
strategy where agents can geometrically transform a model of the scene, based on their own perspective, into a model of the scene 
as seen from the perspective of the hearer. They can then conceptualise the scene from that perspective. 
Communicative success significantly increases and the 
lexicon starts to decrease. However there is ambiguity because the hearer cannot know whether
the speaker has conceptualised the 
scene from his own viewpoint or performed a perspective reversal, leading to a large number of failed communications 
(more than 50 \%). 
\item In stage \textsc{iii}, a more elaborate strategy is used. The speaker now expresses whether he has performed perspective 
reversal (as in `{\itshape to your} left' vs. simply `left'). Only with this strategy are agents able to reach high 
communicative success with a stable inventory. 
\end{itemize}

@@

The earliest attempts to investigate the emergence of syntax date back to 1997, in other words even before 
the first Talking Heads experiment started. The first report 
appeared as a paper in the Artificial Intelligence journal as \citet{Steels:1998}. 
It was described by the journal editors (Dan Bobrow and 
Mike Brady) as ``obviously in an early stage, but very provocative and suggestive.'' The paper was based on my keynote 
lecture at \textsc{ijcai} 1997 in Nagoya and used a precursor of the pan-tilt cameras built by Tony Belpaeme. 
(see \figref{f:plate7}).

@@

\begin{figure}[t]
\centerline{
\includegraphics[width=0.75\textwidth]{chap10/figs/associations.pdf}}
\caption{\label{fig:associations}Top: Two lexical associations have become active (association-259 and association-234) to cover the distinctive 
feature-set ([VISIB v-411] [VER v-431]) resulting from discrimination and perception. Bottom: 
The group of words and the meanings match with schemas forming part of a grammatical association that combines the words. }
\end{figure}
An application of these schemas in producing the utterance `(W O) (W U)' is represented in \figref{fig:associations} and shown 
here: 
\begin{verbatim}
127 ++> Speaker: head-40 Topic: 2 Context: 6 5 4 3 1 0
Categorial Perception:
  ([VISIB v-4111] [VER v-431])
  ([VER v-4311] [\textsc{area} v-4061])
Lexicon lookup: (Association-259 Association-234)
Syntactic structure:
  (syn-cat-77
    (syn-slot-50 (syn-cat-75 |(W O)|))
    (syn-slot-51 (syr-cat-76 |(W U)|)))
Semantic structure:
  (sem-cat-78
    (sem-slot-50 (sem-cat-75 (VER v-431)))
    (sem-slot-51 (sem-cat-76 (VISIB v-411))))
Meaning:
  ([VISIB v-4111] [VER v-4311)
Expression: ((W 0) (W U))
\end{verbatim}

@@

\subsection{The Case Grammar Experiments}

@@

Based on these components, I implemented an initial experiment with two agents.\footnote{These experiments were first reported at the Harvard Evolution of Language conference in 2002 with a paper by Luc Steels
entitled ``Computer Simulations of the origins of Case grammar''. The paper was later deemed 
``too incomprehensible for the Evolution of Language community'' to be acceptable for the conference 
proceedings edited by Maggie Tallerman.}
In 2005 Remi van Trijp picked up this line of research again and worked out the experiment 
further with great skill and sophistication. He extended the 
2-agent simulation to a multi-agent simulation and could finally report a complete running system in his 2008 Ph.D 
thesis \citep{VanTrijp:2014}. Here I give only a brief account of the first attempts towards case grammar
and how this pushed the state of the art in language game research.

But first we need a clearer idea on what case grammar is.
In the sentence `John gave Mary a book', John plays the role of agent, the giver 
of the book, Mary that of recipient or 
beneficiary and the book is the ``patient'' or object of the action. Languages differ in terms of the strategies that they 
use to express these participant roles. Some of them, like German, have an elaborate system of morphological 
markers attached to constituents of the noun phrase. These markers signal that the 
phrase is nominative, accusative, dative, etc., and these cases then map in complex ways 
onto semantic roles such as agent and patient.
Other languages, like English, use a strategy based on word order and prepositions. For example, 
`Mary gave John a book' contains the same participants as `John gave Mary a book' but now Mary is the 
agent and John the recipient. Still other languages, like Japanese, use a strategy based on particles following 
the noun, as in the Japanese sentence `Tanaka-san wa Tokyo de o-to-san ni atta' 
meaning Tanaka met his father in Tokyo. Tanaka-san is the name of a person, oto-san means `father' and atta 
means `met'. The markers are `wa' for the topic, `de' for location and `ni' for another sense of location which can 
also function as possessive. The case grammar experiments focused on modeling the latter strategy.

It used the comparative method again to compare three progressively more complex strategies: 
\begin{enumerate}
\item {\bfshape  Lexical strategy}. The experiment starts from a purely 
lexical language that is predefined in all agents. The lexicon contained words for expressing properties of 
objects (such as `smooth', `red', etc.)
as well as actions involving objects (such as `move-away-from', `move-towards' or `push'). English words
were chosen to make the experiment easier to follow. Agents form
multi-word utterances using straightforward lexicon lookup, producing utterances
like `move-towards red jill' (Jill moves towards the red object) or `push 
jack jill green' (Jack pushes the green block to Jill). There is no syntax at all. Agents can nevertheless already 
achieve communicative success when they use the real world as a way to figure out the participant roles, even though the 
utterance does not provide any information. For example, to detect whether Jack pushes the green block to Jill or Jill 
pushes the green block to Jack.

\item {\bfshape  Specific marker strategy}. The second strategy introduces markers in the form of particles added to words referring 
to objects, in order to specify what argument the object fills in the predicate describing the 
event. For example, there could be two markers: `pi' 
and `pa' specifying that the referent plays the role respectively of giver and gift in a give event, as in
`pi jill pa block give'. The markers are entirely specific for each event in the sense that there is no generalisation 
yet to 'agent', 'patient' or other more abstract semantic roles. This already yields a more effective 
communication system because the complexity of seeking an interpretation matching with the world model is heavily
reduced and residual ambiguity (i.e. there is still more than one way in which the 
meaning matches with the world model) is also avoided.

@@

\subsubsection{Object identification} The first subsystem attempts to detect and track visual units at different hierarchical levels. 
It detects and 'latches onto' regions in the image that are generated by objects of 
interest in the environment. This results in deictic pointers or ``anchors'' that establish and monitor indexical references 
between internal symbols and external objects. Tracking not only takes place for a single object, but for 
an open-ended set of objects at different hierarchical levels, as long as they are part of the same spatio-temporal context.

@@

\textsc{cogar} represents syntactic and semantic structures as frames with units, slots and 
constraints on the slots. The frames are in many ways similar to the feature structures used in feature-structure 
based grammars such as \textsc{hpsg}, although there are many important differences. For example, each unit has a 
unique name so that it becomes possible to refer back to units in defining constraints. \textsc{cogar} also used 
logic variables (denoted by a symbol with a question mark in front) as opposed to path descriptions, which 
lead to the adoption of techniques from logic programming. 
The syntactic and semantic structure of a particular utterance are coupled together as a coupled feature structure 
by making the unit-names shared. Often both are combined when the coupled feature structure needs to be displayed
as in this example below. 
\begin{verbatim}
((unit-15
   (scope utterance)
   (subunits (unit-17 unit-18 unit-16))
   (ordering (unit-17 unit-18 unit-16)))
 (unit-17
   (form (smooth))
   (meaning ((smooth ?source-object1)))
   (referent ?source-object1) 
   (gramcat (takes-marker-1)))
 (unit-18 (form (po)))
 (unit-16 
   (form (move-away-from)) 
   (meaning
     ((move-away-from ?event ?state) 
      (move-away-from-1 ?event ?object)
      (move-away-from-2 ?event ?source-object1)))))
\end{verbatim}
This example could have been constructed, either in parsing or producing, for 
the utterance `po smooth move-away-from'. Syntactic properties such as the form of the word, 
the ordering of the units, or the grammatical categories of a unit are represented explicitly on the syntactic side and 
the meaning and referent are represented on the semantic side. In general, \textsc{cogar} allows the representation of
any kind of information at whatever level of linguistic analysis as part of a feature structure.

@@

Here is an example of a lexical rule that defines the meaning of the word for an event `move-away-from'. The referent is 
the object that is referred to by the word, in this case it is the event itself. 
\clearpage
\begin{verbatim}
((?unit 
   (meaning
      (== (move-away-from ?event ?state)
          (move-away-from-1 ?event ?object)
          (move-away-from-2 ?event ?source)))
   (referent ?event)) 
<-->
((?unit
   (form (== move-away-from)))))
\end{verbatim}
Next, here is an example of a rule for the marker `po'. On the semantic side it ensures that the referent of one unit 
whose referent is ?source is equal to the object filling the argument move-away-from-2 in a move-away-from event. 
On the syntactic side, it triggers when there is a word that has the grammatical category `takes-marker-1' and 
is preceded by the marker `po'. (<< stands for `precedes') 
\begin{verbatim}
((?predicate 
   (meaning
    (== (move-away-from ?evnt)
        (move-away-from-1 ?evnt ?patient)
        (move-away-from-2 ?evnt ?source))))
 (?argument
   (referent ?source))) 
<---> 
((?argument
   (gramcat (== takes-marker-1)))
 (?marker-unit
   (form (== po))) 
(<< ?marker-unit ?argument))))
\end{verbatim}

There are additional rules that specify which word takes which marker, by assigning them to a particular 
grammatical category. For example, the word `smooth' has the category `takes-marker-1': 
\begin{verbatim}
((?unit
   (gramcat (== takes-marker-1)) 
   (form (== smooth))))
<---> 
((?unit (form (== smooth))))
\end{verbatim}

The marker `po' in the above example is entirely specific for one argument of the move-away-from predicate, as would be the 
case when the second strategy is enacted. To implement the third strategy 
(to achieve more abstract semantic roles) requires that the markers become generic, which implies that 
we get additional rules that recategorise the arguments of a predicate in terms of 
semantic roles like: agent, patient, source, etc. Here is an example of a rule that 
has this effect. role-1 is an example of a semantic role created by the agent as an abstraction of the 
move-away-from-2 argument. 
\begin{verbatim}
((?predicate
   (meaning
    (== (move-away-from ?evnt) 
        (move-away-from-1 ?evnt ?patient)
        (move-away-from-2 ?evnt ?source)))))
<---> 
 ((?predicate
    (meaning
      (== (move-away-from ?evnt)
          (move-away-from-1 ?evnt ?patient)
          (move-away-from-2 ?evnt ?source)))) 
    (expanded-meaning (== (role-1 ?evnt ?source))))
\end{verbatim}
And the rules that click everything together now operate over semantic roles as shown in the following example: 
\begin{verbatim}
 ((?predicate
   (expanded-meaning (== (role-1 ?event ?source))))
  (?unit (referent ?source))) 
<---> 
 ((?unit (gramcat (== takes-marker-1)))
  (?marker-unit (form (== po))) 
 (<< ?marker-unit ?unit))
\end{verbatim}
Recall that all rules must be reversible and that the conditional part must match and the concluding part must unify.

>>>>

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.50\textwidth]{chap10/figs/look-ball.pdf}
\includegraphics[width=.45\textwidth]{chap10/figs/views-ball.pdf}}
\caption{\label{f:look-ball}Left: Fr\'{e}d\'{e}ric Kaplan interacts with the \textsc{aibo} robot in an experiment for the social learning of lexicons.
Right: View of the ball through the eye of the robot. As the same object is now seen from many different viewpoints it 
becomes much more challenging to consistently recognise it.}
\end{figure}
The first \textsc{aibo} experiments did not involve a population of agents but rather a single agent that was interacting directly with a human
experimenter. The agent had to learn the lexicon of the human through social 
interaction within the context of situated language games. The 2001
paper \dblenquote{\textsc{aibo}'s first words} by Steels and Kaplan \nocite{Steels:01h} described experiments in which an \textsc{aibo} robot 
learned words about objects in the environment, such as \smplenquote{ball} or \smplenquote{smiley}, as well as words for actions, such as 
\smplenquote{sit down} or \smplenquote{stand up}.

A typical example of a game is the DO game, used to learn the names of behaviours.\is{DO game} 
The robot is programmed with a repertoire of behaviours, such as 
walk, sit down, stand up, look left/right/up/down, push ball, etc. When the human utters a word, the hearing robot 
looks up the word in its own memory and performs that behaviour. If it does not know the word, it 
performs one of the behaviours for which it does not have a word yet. 
If this behaviour is the wrong one, the robot receives feedback and correction, 
thus learning (i) that the behaviour is not associated with the word and (ii) what the right word is for the behaviour 
just performed. When the right behaviour was chosen, the association between this behaviour 
and the word can be stored if the association was not yet in memory, or reinforced if it was using the 
lateral inhibition learning strategy. 
Games therefore always come with two variant scripts. There is one with a successful interaction, which then 
leads to a reinforcement of the existing lexicon, and one with a non-successful interaction, where the 
robot gets corrected and learns a new word. Here are the two variants for the DO Game: 
\begin{center}
\begin{tabular}{ l  l }
\lsptoprule
{\itshape Reinforcement script}&{\itshape Correction script} \\ \midrule
Human: Listen, Walk. & H: Listen, Walk.  \\ 
R: Walk? & R: Walk? \\
 {\itshape (walks)} & {\itshape (sits)} \\
H: Yes & H: No, this is \smplenquote{sit}. \\
& R: Sit ? \\ 
& H: Yes. \\ 
\lspbottomrule
\end{tabular}
\end{center}
The scripts of the agents are represented internally using probabilistic finite state machines with 
temporal annotations. This computational formalism (pioneered by Rodney Brooks) was also used to program the 
physical behaviour of the robots (\figref{fig:fsm}).\footnote{
This architecture was first described in \cite{Brooks:1986}. 
For the experiments I implemented a LISP-based Behaviour Language that ran also on the \textsc{aibo}.}

@@

The \textsc{aibo} experiments were the first robotic experiments demonstrating social, situated learning in action. 
They were highly relevant for discovering the subtle interaction patterns that would work and 
could be realistically implemented in human-robot scenarios. The goal of these experiments was 
not to address the question of the origins of a 
new language in a population of agents. However, another experiment, still with the \textsc{aibo} robots,  
did substantially advance the state of the art: the Perspective Reversal experiment carried out by Martin Loetzsch 
and Luc Steels in 2004--2005.\footnote{
Very little was published about this experiment, partly because reviewers had the greatest difficulty to grasp 
its underlying rationale or understand the technical details. The main paper is \cite{Steels:2008spatial}. 
Subsequent work by Michael Spranger significantly scaled up the experiment, as discussed in the next chapter.}

\subsection{The Perspective Reversal experiment}

The Perspective Reversal experiment \is{Perspective Reversal} focused on the question how a spatial language could emerge that employed
perspective reversal. It was the first experiment to systematically compare different language strategies, thus 
introducing for the first time the {\itshape comparative method} \is{comparative method} 
that would inform all later language game experiments.

@@

The Perspective Reversal experiment applied this methodology to explain a remarkable universal feature 
of human languages, namely, 
that speakers may use a different perspective on the scene than their own when 
conceptualising what to say, and that they may explicitly mark perspective switching, using lexical or grammatical means. Suppose
a speaker is facing a hearer, then it indeed makes a big difference whether she says 
\emph{the door to your left}  or \emph{the door to my left}. If the speaker says
\emph{the door to my left} she expects the hearer to perform an egocentric
perspective transformation and see the situation from her own point of view. Perspective switching is here signalled by 
explicitly expressing which perspective is used: {\itshape your} left versus {\itshape my} left. 
The German prepositions \emph{herein} and \emph{hinein}, both meaning \smplenquote{into} are another example of perspective marking. They
distinguish whether the direction of movement is 
towards the speaker \emph{herein} or away from the speaker \emph{hinein}, analogous to English \emph{come} (to where I am) 
versus \emph{go} (to another location). Although there are obvious differences in how languages express
perspective, there can be no doubt that perspective marking is pervasive across 
languages and language families, and that speakers make abundant use of it.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.65\textwidth]{chap10/figs/tokyo-persp.pdf}}
\caption{\label{fig:tokyo-persp}Luc Steels demonstrates live the Perspective Reversal experiment in Tokyo, April 2005. \textsc{aibo} robots track 
the movement of a ball and 
the robot acting as speaker describes to the hearer an observed movement, possibly taking into account the 
perspective of the hearer. For example, the speaker might say \smplenquote{the ball rolls from my left to your right}.}
\end{figure}

\enlargethispage{1\baselineskip}
The Perspective Reversal experiment worked as follows. A population of \textsc{aibo} robots roam around
freely in an unconstrained in-door environment. As soon as one sees the
ball, it comes to a stop and searches for another robot nearby, which also
looks for the ball and stops when it sees it. Then the human
experimenter pushes the ball with a stick so that it rolls a short
distance (see \figref{fig:tokyo-persp}). The two robots next play a description game. One robot (acting as the \smplenquote{speaker}) 
describes the ball-moving event to the other robot (the \dblenquote{hearer}), going through the same semiotic cycle as the Talking 
Heads agents. The game is a success if the meaning conveyed by the speaker is compatible with the hearer's 
own perception of the scene.

@@

Within this experimental setting three different strategies were tried in a long-term experiment where
a new strategy was injected at certain stages and performance monitored: 
\begin{itemize}
\item In stage \textsc{i}, agents use 
the Discrimination Game strategy and the Naming Game strategy discussed in chapters 4 and 5. A 
lexicon is emerging and there are some successful games (those for which the speaker and the hearer see the situation from 
the same perspective). But the lexicon keeps expanding and success is very limited as games where robots have a different 
perspective on the rolling ball all fail. 
\item In stage \textsc{ii}, agents have an extended 
strategy where agents can geometrically transform a model of the scene, based on their own perspective, into a model of the scene 
as seen from the perspective of the hearer. They can then conceptualise the scene from that perspective. 
Communicative success significantly increases and the 
lexicon starts to decrease. However there is ambiguity because the hearer cannot know whether
the speaker has conceptualised the 
scene from his own viewpoint or performed a perspective reversal, leading to a large number of failed communications 
(more than 50 \%). 
\item In stage \textsc{iii}, a more elaborate strategy is used. The speaker now expresses whether he has performed perspective 
reversal (as in \smplenquote{{\itshape to your} left} vs. simply \smplenquote{left}). Only with this strategy are agents able to reach high 
communicative success with a stable inventory. 
\end{itemize}

@@

The earliest attempts to investigate the emergence of syntax date back to 1997, in other words even before 
the first Talking Heads experiment started. The first report 
appeared as a paper in the Artificial Intelligence journal as \citet{Steels:1998}. 
It was described by the journal editors (Dan Bobrow and 
Mike Brady) as \dblenquote{obviously in an early stage, but very provocative and suggestive.} The paper was based on my keynote 
lecture at \textsc{ijcai} 1997 in Nagoya and used a precursor of the pan-tilt cameras built by Tony Belpaeme. 
(see \figref{f:plate7}).

@@

\begin{figure}[t]
\centerline{
\includegraphics[width=0.75\textwidth]{chap10/figs/associations.pdf}}
\caption{\label{fig:associations}Top: Two lexical associations have become active (association-259 and association-234) to cover the distinctive 
feature-set ([VISIB v-411] [VER v-431]) resulting from discrimination and perception. Bottom: 
The group of words and the meanings match with schemas forming part of a grammatical association that combines the words. }
\end{figure}
An application of these schemas in producing the utterance \smplenquote{(W O) (W U)} is represented in \figref{fig:associations} and shown 
here: 
\begin{verbatim}
127 ++> Speaker: head-40 Topic: 2 Context: 6 5 4 3 1 0
Categorial Perception:
  ([VISIB v-4111] [VER v-431])
  ([VER v-4311] [\textsc{area} v-4061])
Lexicon lookup: (Association-259 Association-234)
Syntactic structure:
  (syn-cat-77
    (syn-slot-50 (syn-cat-75 |(W O)|))
    (syn-slot-51 (syr-cat-76 |(W U)|)))
Semantic structure:
  (sem-cat-78
    (sem-slot-50 (sem-cat-75 (VER v-431)))
    (sem-slot-51 (sem-cat-76 (VISIB v-411))))
Meaning:
  ([VISIB v-4111] [VER v-4311)
Expression: ((W 0) (W U))
\end{verbatim}

@@

\subsection{The Case Grammar experiments}

@@

Based on these components, I implemented an initial experiment with two agents.\footnote{These experiments were first reported at the Harvard Evolution of Language conference in 2002 with a paper by Luc Steels
entitled \dblenquote{Computer Simulations of the origins of Case grammar}. The paper was later deemed 
\dblenquote{too incomprehensible for the Evolution of Language community} to be acceptable for the conference 
proceedings edited by Maggie Tallerman.}
In 2005 Remi van Trijp picked up this line of research again and worked out the experiment 
further with great skill and sophistication. He extended the 
2-agent simulation to a multi-agent simulation and could finally report a complete running system in his 2008 Ph.D 
thesis \citep{VanTrijp:2014}. Here I give only a brief account of the first attempts towards case grammar
and how this pushed the state of the art in language game research.

But first we need a clearer idea on what case grammar is.
In the sentence \smplenquote{John gave Mary a book}, John plays the role of agent, the giver 
of the book, Mary that of recipient or 
beneficiary and the book is the \dblenquote{patient} or object of the action. Languages differ in terms of the strategies that they 
use to express these participant roles. Some of them, like German, have an elaborate system of morphological 
markers attached to constituents of the noun phrase. These markers signal that the 
phrase is nominative, accusative, dative, etc., and these cases then map in complex ways 
onto semantic roles such as agent and patient.
Other languages, like English, use a strategy based on word order and prepositions. For example, 
\smplenquote{Mary gave John a book} contains the same participants as \smplenquote{John gave Mary a book} but now Mary is the 
agent and John the recipient. Still other languages, like Japanese, use a strategy based on particles following 
the noun, as in the Japanese sentence \smplenquote{Tanaka-san wa Tokyo de o-to-san ni atta} 
meaning Tanaka met his father in Tokyo. Tanaka-san is the name of a person, oto-san means \smplenquote{father} and atta 
means \smplenquote{met}. The markers are \smplenquote{wa} for the topic, \smplenquote{de} for location and \smplenquote{ni} for another sense of location which can 
also function as possessive. The case grammar experiments focused on modeling the latter strategy.

It used the comparative method again to compare three progressively more complex strategies: 
\begin{enumerate}
\item {\bfshape  Lexical strategy}. The experiment starts from a purely 
lexical language that is predefined in all agents. The lexicon contained words for expressing properties of 
objects (such as \smplenquote{smooth}, \smplenquote{red}, etc.)
as well as actions involving objects (such as \smplenquote{move-away-from}, \smplenquote{move-towards} or \smplenquote{push}). English words
were chosen to make the experiment easier to follow. Agents form
multi-word utterances using straightforward lexicon lookup, producing utterances
like \smplenquote{move-towards red jill} (Jill moves towards the red object) or `push 
jack jill green' (Jack pushes the green block to Jill). There is no syntax at all. Agents can nevertheless already 
achieve communicative success when they use the real world as a way to figure out the participant roles, even though the 
utterance does not provide any information. For example, to detect whether Jack pushes the green block to Jill or Jill 
pushes the green block to Jack.

\item {\bfshape  Specific marker strategy}. The second strategy introduces markers in the form of particles added to words referring 
to objects, in order to specify what argument the object fills in the predicate describing the 
event. For example, there could be two markers: \smplenquote{pi} 
and \smplenquote{pa} specifying that the referent plays the role respectively of giver and gift in a give event, as in
\smplenquote{pi jill pa block give}. The markers are entirely specific for each event in the sense that there is no generalisation 
yet to 'agent', 'patient' or other more abstract semantic roles. This already yields a more effective 
communication system because the complexity of seeking an interpretation matching with the world model is heavily
reduced and residual ambiguity (i.e. there is still more than one way in which the 
meaning matches with the world model) is also avoided.

@@

\subsubsection{Object identification} The first subsystem attempts to detect and track visual units at different hierarchical levels. 
It detects and 'latches onto' regions in the image that are generated by objects of 
interest in the environment. This results in deictic pointers or \dblenquote{anchors} that establish and monitor indexical references 
between internal symbols and external objects. Tracking not only takes place for a single object, but for 
an open-ended set of objects at different hierarchical levels, as long as they are part of the same spatio-temporal context.

@@

\textsc{cogar} represents syntactic and semantic structures as frames with units, slots and 
constraints on the slots. The frames are in many ways similar to the feature structures used in feature-structure 
based grammars such as \textsc{hpsg}, although there are many important differences. For example, each unit has a 
unique name so that it becomes possible to refer back to units in defining constraints. \textsc{cogar} also used 
logic variables (denoted by a symbol with a question mark in front) as opposed to path descriptions, which 
lead to the adoption of techniques from logic programming. 
The syntactic and semantic structure of a particular utterance are coupled together as a coupled feature structure 
by making the unit-names shared. Often both are combined when the coupled feature structure needs to be displayed
as in this example below. 
\begin{verbatim}
((unit-15
   (scope utterance)
   (subunits (unit-17 unit-18 unit-16))
   (ordering (unit-17 unit-18 unit-16)))
 (unit-17
   (form (smooth))
   (meaning ((smooth ?source-object1)))
   (referent ?source-object1) 
   (gramcat (takes-marker-1)))
 (unit-18 (form (po)))
 (unit-16 
   (form (move-away-from)) 
   (meaning
     ((move-away-from ?event ?state) 
      (move-away-from-1 ?event ?object)
      (move-away-from-2 ?event ?source-object1)))))
\end{verbatim}
This example could have been constructed, either in parsing or producing, for 
the utterance \smplenquote{po smooth move-away-from}. Syntactic properties such as the form of the word, 
the ordering of the units, or the grammatical categories of a unit are represented explicitly on the syntactic side and 
the meaning and referent are represented on the semantic side. In general, \textsc{cogar} allows the representation of
any kind of information at whatever level of linguistic analysis as part of a feature structure.

@@

Here is an example of a lexical rule that defines the meaning of the word for an event \smplenquote{move-away-from}. The referent is 
the object that is referred to by the word, in this case it is the event itself. 
\clearpage
\begin{verbatim}
((?unit 
   (meaning
      (== (move-away-from ?event ?state)
          (move-away-from-1 ?event ?object)
          (move-away-from-2 ?event ?source)))
   (referent ?event)) 
<-->
((?unit
   (form (== move-away-from)))))
\end{verbatim}
Next, here is an example of a rule for the marker \smplenquote{po}. On the semantic side it ensures that the referent of one unit 
whose referent is ?source is equal to the object filling the argument move-away-from-2 in a move-away-from event. 
On the syntactic side, it triggers when there is a word that has the grammatical category \smplenquote{takes-marker-1} and 
is preceded by the marker \smplenquote{po}. (<< stands for \smplenquote{precedes}) 
\begin{verbatim}
((?predicate 
   (meaning
    (== (move-away-from ?evnt)
        (move-away-from-1 ?evnt ?patient)
        (move-away-from-2 ?evnt ?source))))
 (?argument
   (referent ?source))) 
<---> 
((?argument
   (gramcat (== takes-marker-1)))
 (?marker-unit
   (form (== po))) 
(<< ?marker-unit ?argument))))
\end{verbatim}

There are additional rules that specify which word takes which marker, by assigning them to a particular 
grammatical category. For example, the word \smplenquote{smooth} has the category \smplenquote{takes-marker-1}: 
\begin{verbatim}
((?unit
   (gramcat (== takes-marker-1)) 
   (form (== smooth))))
<---> 
((?unit (form (== smooth))))
\end{verbatim}

The marker \smplenquote{po} in the above example is entirely specific for one argument of the move-away-from predicate, as would be the 
case when the second strategy is enacted. To implement the third strategy 
(to achieve more abstract semantic roles) requires that the markers become generic, which implies that 
we get additional rules that recategorise the arguments of a predicate in terms of 
semantic roles like: agent, patient, source, etc. Here is an example of a rule that 
has this effect. role-1 is an example of a semantic role created by the agent as an abstraction of the 
move-away-from-2 argument. 
\begin{verbatim}
((?predicate
   (meaning
    (== (move-away-from ?evnt) 
        (move-away-from-1 ?evnt ?patient)
        (move-away-from-2 ?evnt ?source)))))
<---> 
 ((?predicate
    (meaning
      (== (move-away-from ?evnt)
          (move-away-from-1 ?evnt ?patient)
          (move-away-from-2 ?evnt ?source)))) 
    (expanded-meaning (== (role-1 ?evnt ?source))))
\end{verbatim}
And the rules that click everything together now operate over semantic roles as shown in the following example: 
\begin{verbatim}
 ((?predicate
   (expanded-meaning (== (role-1 ?event ?source))))
  (?unit (referent ?source))) 
<---> 
 ((?unit (gramcat (== takes-marker-1)))
  (?marker-unit (form (== po))) 
 (<< ?marker-unit ?unit))
\end{verbatim}
Recall that all rules must be reversible and that the conditional part must match and the concluding part must unify.
